---
title: "NZ Covid Vaccination"
output:
  html_document:
    df_print: paged
---

```{r set_up, include=FALSE, eval=TRUE}
rm(list = ls()); gc()

library(tidyverse)
library(data.table)
library(readxl)
library(broom)
library(gridExtra)
library(lemon)
library(scales)
library(ggrepel)

setwd("/Users/danielpringle/Code/covidvaxnz/R")
data_dir <- file.path("/Users/danielpringle/Code/covidvaxnz/data")
  
```

## Ministry of Health Data

Data downloaded from [NZ Ministry of Health](https://www.health.govt.nz/our-work/diseases-and-conditions/covid-19-novel-coronavirus/covid-19-data-and-statistics/covid-19-vaccine-data#download) 

### Reading Data

```{r read_prepare_data, eval=TRUE, results='hide', echo=FALSE,  message=FALSE}
# https://www.health.govt.nz/system/files/documents/pages/211024_-_cvip_equity_-_rate_ratios_and_uptake_over_time.xlsx

#file <- "/Users/danielpringle/Code/covidvaxnz/data/211024_cvip_equity.xlsx"

# Update MOH file and download 
url.base <- "https://www.health.govt.nz/system/files/documents/pages/"
url.file <- "211107_-_cvip_equity_-_rate_ratios_and_uptake_over_time.xlsx"
maxdate <- as.Date(substr(url.file, 1, 6), format = "%y%m%d")

url <- paste0(url.base, url.file)
data.file <- file.path(data_dir, "latest_cvip_equity.xlsx")
download.file(url, data.file)

sheets <- data.file %>%
  excel_sheets() %>% 
  set_names() 

class(sheets)

x <- lapply(sheets, function(X) readxl::read_excel(data.file, sheet = X))
x <- lapply(x, as.data.table)
names(x) <- sheets

# Long table of vaccination data, and HSU population data by age-ethnic splits
vx_query <- x[["Vaccination Query"]]
dhb_pop <- x[["HSU Table"]]

# Join data and tidy up
vx <- merge(x = vx_query, y = dhb_pop, by = c("Ethnic group", "Age group", "DHB of residence"))
oldnames = c("Week ending date", 
             "Dose number", 
             "Ethnic group", 
             "Age group", 
             "DHB of residence",
             "# doses administered", 
             "# people (HSU)")
newnames = c("week", 
             "dose", 
             "ethnicgroup", 
             "agegroup", 
             "DHB", 
             "count", 
             "popsegment")
setnames(vx, old = oldnames, new = newnames)
setcolorder(vx, newnames)

# Set as date, remove incomplete data sometimes present for next week
vx$week <- as.Date(vx$week)
vx <- vx[week <= maxdate]

# Set as factors
vx$dose <- factor(vx$dose)
vx$agegroup <- factor(vx$agegroup)
vx$ethnicgroup <- factor(vx$ethnicgroup)

# Remove 'Unknown', 'Overseas' and 'Overseas and undefined'
vx <- vx[!get("DHB") %in% c('Unknown', 'Overseas','Overseas and undefined')] 

vx$DHB <- factor(vx$DHB)

# pop as HSU population stats with above names
pop <- dhb_pop
oldnames = c("Ethnic group", 
             "Age group", 
             "DHB of residence",
             "# people (HSU)")
newnames = c("ethnicgroup", 
             "agegroup", 
             "DHB",
             "popsegment")
setnames(pop, old = oldnames, new = newnames)
setcolorder(pop, newnames)
setorderv(pop, c( "DHB", "ethnicgroup", "agegroup"))

pop <- pop[!get("DHB") %in% c('Unknown', 'Overseas','Overseas and undefined')] 
pop$DHB <- factor(pop$DHB)
pop$agegroup <- factor(pop$agegroup)
pop$ethnicgroup <- factor(pop$ethnicgroup)
```

### Data contains weekly vaccination counts and population demographics by DHB 

Data includes counts of first and second dose vaccinations, by week, split by Ethnic group, Age group, and DHB oof residence. It also includes a separate table of HSU population statistics, split into the same groups for each DHB. These numebrs are used to determine percentage vaccination rates, cut in various ways.

```{r inspect_data, include=TRUE, asis = TRUE}
str(vx)
head(vx, 20)
str(pop)
head(pop, 20)
```

## Summary Statistics

Before looking at vaccination rates, split out in detail, let's summarise the overall data:

```{r pop_by_DHB, include=TRUE, echo = FALSE}
# Total population by DHB
pop_dhb_tot <- pop[ , .(population = sum(popsegment), ethnicgroup = "Total"), by = c("DHB")]
pop_dhb_eth <- pop[ , .(population = sum(popsegment)), by = c("DHB", "ethnicgroup")]
pop_dhb <- rbind(pop_dhb_tot, pop_dhb_eth)

p_pop_tot <- ggplot(data = pop_dhb[ethnicgroup == "Total"], aes(x = reorder(DHB, population), y = population/1000)) +
  geom_col(fill = "lightblue") +
  coord_flip() +
  labs(y = "Population (1000s)",
       x = "") +
  ggtitle("Total Population by DHB") +
  theme_minimal()
grid.arrange(p_pop_tot)
```

### Ethnic Group

MOH report data by Ethnic group and age group, and both factors have been commented on in the media. The chart belows below show the Ethnic group split reported by DHB, ordered by total population.

```{r pop_by_DHB_eth, include=TRUE, echo = FALSE}
# Total population by DHB and ethnic group
pop_dhb_tot <- pop[ , .(population = sum(popsegment), ethnicgroup = "Total"), by = c("DHB")]
pop_dhb_eth <- pop[ , .(population = sum(popsegment)), by = c("DHB", "ethnicgroup")]
pop_dhb <- rbind(pop_dhb_tot, pop_dhb_eth)

p_pop_tot_eth <- ggplot(data = pop_dhb, aes(x = reorder(DHB, population), y = population/1000)) +
  geom_col(aes(fill = ethnicgroup)) +
  coord_flip() +
  labs(y = "Population (1000s)",
       x = "") +
  scale_y_continuous(limits = c(0,700),
                     breaks = seq(0,600,200)) +
  ggtitle("Population by Ethnic Group by DHB") +
  geom_text(aes(label = round(population/1000)), hjust = -0.2, colour = "darkgrey") +
  facet_wrap(~ethnicgroup, nrow = 1) +
  theme_minimal() +
  theme(legend.position='none')

grid.arrange(p_pop_tot_eth, nrow = 1)
```

```{r frac_pop_by_DHB_eth, include=TRUE, echo = FALSE}
# Total population by DHB and ethnic group

frac_pop_dhb_eth <- pop[ , .(population = sum(popsegment)), 
                by = c("DHB", "ethnicgroup")]
frac_pop_dhb_eth <- frac_pop_dhb_eth[ , `:=`(fraction = round(population/sum(population),3)),
                by = c("DHB")]

p_frac_pop_eth <- ggplot(data = frac_pop_dhb_eth[ethnicgroup %in% c("Māori", "Pacific")], aes(x = reorder(DHB, fraction), y = fraction)) +
  geom_col(aes(fill = ethnicgroup)) +
  coord_flip() +
  labs(y = "Fraction of population by DHB",
       x = "") +
  scale_y_continuous(label = scales::percent_format(accuracy = 1),
                     limits = c(0,0.5),
                     breaks = seq(0.1,0.5,0.1)) +
  ggtitle("Fraction of  Māori and Pacific population by DHB") +
  theme_minimal() +
  facet_wrap(~ethnicgroup)+
  theme(legend.position='none')

grid.arrange(p_frac_pop_eth, nrow = 1)

```

### Age Group

The vaccination target is two doses of pfizer vaccine for 90% of the eligible population in each DHB. At present, the ineligible popualtion is people with medical exemptions and children under the age of 12. The population data accessed here does not include ondere 12s. That is, the sum for each DHB over all age and ethnic groups is the vaccine-eligible populaiton (apart from small numbers of exemptions).

Another age realted factos is the initial roll out to more vulnerable people, including those over 65 years of age. This should be kept in mind when comoparing vaccination uptake, as some DHBs have older (or younger) poplautions than others. 

MOH report data by Ethnic group and age group, and both factors have been commented on in the media. The chart belows below show the Ethnic group split reported by DHB, againi ordered by total population.

```{r pop_by_DHB_age, include=TRUE, echo = FALSE}
# Total population by DHB and ethnic group
pop_dhb_age <- pop[ , .(population = sum(popsegment)), 
                by = c("DHB", "agegroup")]
pop_dhb_age <- pop_dhb_age[ , `:=`(fraction = round(population/sum(population),3)),
                by = c("DHB")]

p_age_frac65 <- ggplot(data = pop_dhb_age[agegroup=="65+"], aes(x = reorder(DHB, fraction), y = fraction)) +
    geom_col(fill = "lightblue") +
  coord_flip() +
  labs(y = "Fraction over 65 years old",
       x = "") +
  scale_y_continuous(label = scales::percent_format(accuracy = 1),
                     limits = c(0,0.35),
                     breaks = seq(0.1,0.35,0.05)) +
  ggtitle("Fration of Population 65+") +
  theme_minimal() 

grid.arrange(p_age_frac65, nrow = 1)

```


## Uptake Rates

### Overall NZ Uptake Rate

Let's now look at the vaccination uptake rates, recorded as first or second dose counts per week. 

```{r vax_uptake_nz,  include = TRUE, echo = FALSE, comment ="",, fig.align='center'}
tail(vx)
head(pop)
pop_nz_eligible = sum(pop$popsegment)
names(vx)
v_nz <- vx[ , .(count = sum(count),
                pop = pop_nz_eligible), .(week, dose)]

setorderv(v_nz, c("week", "dose"))
v_nz[ , `:=`(total = cumsum(count)), .(dose)]
v_nz[ , `:=`(uptake = round(total/pop, 3)), .(dose)]

p_v_nz_week <- ggplot(data = v_nz, aes(x = week, y = count, color = dose)) +
  geom_line() +
  scale_x_date() +
  scale_y_continuous(position = "right",label=comma) +
  labs(y = "",
       x = "Week ended") +
  ggtitle("Weekly number of vaccine doses received") +
  theme_minimal() +
  theme(legend.position=c(0.2,0.5), 
        legend.direction = "horizontal")

p_v_nz_total <- ggplot(data = v_nz, aes(x = week, y = total, color = dose)) +
  geom_line() +
  scale_x_date() +
  scale_y_continuous(position = "right",label=comma)+
  labs(y = "",
       x = "") +
  ggtitle("Total number of vaccine doses received to date") +
  theme_minimal() +
  theme(legend.position=c(0.2,0.5), 
        legend.direction = "horizontal")

p_v_nz_uptake <- ggplot(data = v_nz, aes(x = week, y = uptake, color = dose)) +
  geom_line() +
  scale_x_date() +
  scale_y_continuous(position = "right",
                     label = scales::percent_format(accuracy = 1),
                     limits = c(0,1),
                     breaks = seq(0,1,0.2)) +
  labs(y = "",
       x = "") +
  ggtitle("Uptake of vaccine doses (% of eligible population)") +
  theme_minimal() +
  theme(legend.position=c(0.2,0.5), 
        legend.direction = "horizontal")

gA <- ggplotGrob(p_v_nz_week)
gB <- ggplotGrob(p_v_nz_total)
gC <- ggplotGrob(p_v_nz_uptake)
#grid::grid.newpage()
#grid::grid.draw(rbind(gA, gB, gC))
grid.arrange(gA, gB, gC)
```

### Uptake Rate by DHB

Let's now look at the vaccination uptake rates, recorded as first or second dose counts per week. 

```{r vax_uptake_dhb,  include = TRUE, echo = FALSE, comment ="",fig.asp=1, fig.align='center'}
v_dhb <- vx[ , .(count = sum(count)), .(week, dose, DHB)]
pop_dhb_total <- pop_dhb_tot[ethnicgroup == "Total", c("DHB", "population")]

v_dhb <- merge(x = v_dhb, y = pop_dhb_total, by.x = c("DHB"), by.y = c("DHB"), all.x = TRUE)

setorderv(v_dhb, c("DHB", "week", "dose"))
v_dhb[ , `:=`(total = cumsum(count)), .(dose, DHB)]
v_dhb[ , `:=`(uptake = round(total/population, 3)), .(dose, DHB)]

last_data <- v_dhb[get("week") == max(get("week")), 
                   .(week = week,
                     count = count, 
                     total = total, 
                     uptake = uptake ), 
                   .(dose, DHB)]

p_v_dhb_uptake <- ggplot(data = v_dhb, aes(x = week, y = uptake, color = dose)) +
  geom_line() +
  scale_x_date() +
  geom_hline(yintercept = 0.9, linetype = "dashed") +
  scale_y_continuous(position = "right",
                     label = scales::percent_format(accuracy = 1),
                     limits = c(0,1.2),
                     breaks = seq(0,1,0.2)) +
  labs(y = "", x = "") +
  geom_text_repel(aes(label = paste0(round(100*uptake,0), "%")), data = last_data, size = 5) +
  ggtitle(paste0("Uptake of vaccine doses, at ", format(maxdate, "%d %B %Y"))) +
  facet_wrap(~DHB) + 
  theme_bw() +
  # theme(legend.position="bottom", 
  #       legend.direction = "horizontal") +
  theme(legend.position="none") +
  theme(axis.text=element_text(size=10), #change font size of axis text
        axis.title=element_text(size=10), #change font size of axis titles
        plot.title=element_text(size=10), #change font size of plot title
        legend.text=element_text(size=10), #change font size of legend text
        strip.text=element_text(size=10), #change font size of facet strip text
        legend.title=element_text(size=10)) #change font size of legend title   

grid.arrange(p_v_dhb_uptake)

```

## Simple logistic fits

Uptake curves in various situations are modelled mathematically using 'logistic' curves. These are one of a variety of 'S' shaped curves, characterised by a uptake rate with is initially slow, then accelerates and then slows again to level out at some 'saturation' level. Let's see how well logistic curves fit the overall uptake in NZ, and by DHB. If they do fit well, the saturation level may be a useful indicator of how much a gap remains to be made up - throough targeted initiatives - do lift the vacciantion rates, overall and in segments. 

### NZ Overall Uptake

```{r vax_uptake_nz_fit,  include = TRUE, echo = TRUE, comment ="",, fig.align='center'}

p_v_nz_uptake2 <- ggplot(data = v_nz, aes(x = week, y = uptake, color = dose)) +
  geom_line() +
  scale_x_date() +
  scale_y_continuous(position = "right",
                     label = scales::percent_format(accuracy = 1),
                     limits = c(0,1),
                     breaks = seq(0,1,0.2)) +
  labs(y = "",
       x = "") +
  ggtitle("Uptake of vaccine doses (% of eligible population)") +
  theme_minimal() +
  theme(legend.position=c(0.2,0.5), 
        legend.direction = "horizontal")

# m <- glm(surv ~ size, family = binomial, data = v_nz)
# #…and extract fitted values using broom::augment()
# 
# plot_df <- augment(m, type.predict = "response")
# head(plot_df)

grid.arrange(p_v_nz_uptake2)

```





### Correlations between uptake and other factors
